Length Bias and Search Limitations in Cartesian Genetic Programming

### 2013-04-09 Length Bias and Seach Limitations in CGP/main.tex
Added non anonymous version of my paper, fixed Bill's author information
Updated LengthBias to cover reviewer comments, updated bib to cite the paper correctly

\documentclass{sig-alternate}
%\linespread{2.5}
\usepackage{url}
\usepackage{verbatim}
\newcommand{\includegraphicsfit}[1]
{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{GECCO'13,} {July 6-10, 2013, Philadelphia, Pennsylvania, USA.}
\CopyrightYear{2013}
\crdata{TBA}
\clubpenalty=10000
\widowpenalty = 10000
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

\numberofauthors{2} %
\begin{comment}
\author{
% 1st. author
\alignauthor
Brian W. Goldman\\
       \affaddr{BEACON Center for the Study of Evolution in Action}\\
       \affaddr{Michigan State University, U.S.A.}\\
       \email{brianwgoldman@acm.org}
% 2nd. author
\alignauthor
William Punch\\
       \affaddr{BEACON Center for the Study of Evolution in Action}\\
       \affaddr{Michigan State University, U.S.A.}\\
       \email{punch@cse.msu.edu}
}
%\end{comment}
%\begin{comment}
\author{
% 1st. author
\alignauthor
Anonymous\\
       \affaddr{Group}\\
       \affaddr{Organization}\\
       \email{email@site.com}
% 2nd. author
\alignauthor
Anonymous\\
       \affaddr{Group}\\
       \affaddr{Organization}\\
       \email{email@site.com}
}
%\end{comment}

@@

% A category with the (minimum) three required fields
\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

@@

\keywords{Cartesian Genetic Programming, Neutrality, Bloat}
\\
\\
\section{Introduction}
In all black box search, search designers must make assumptions about the properties
of the problem in order to outperform random search.  When these assumptions
match the application, search time can be greatly reduced.  As such, understanding how design decisions
effect search effectiveness or even prevent search from finding solutions is very important.
%
In Cartesian Genetic Programming (CGP) the maximum genome size of an individual
is set at initialization.  While this
requires the maximum genome size to be larger than the problem's solution needs,
the question remains; how will the maximum genome size effect evolution's
speed at finding solutions.  As larger genomes inherently represent larger search
spaces, intuition suggests that using the lowest possible maximum size will have the most efficient search.
However, in CGP a mounting body of evidence suggests the optimal
maximum size is much larger than the actual solutions evolved~\cite{miller:2006:redundancy}.
We examine the cause of this discrepancy by applying CGP and a few
modified versions of CGP to problems which test specific edge cases in representation.
The result is a better understanding of the search limitations of CGP.

@@

For all problems except \emph{Neutral} (Section~\ref{sec-neutral}), we provide a plot of the median
evaluations until success using different problem sizes.  If the median run
was unsuccessful, no information is plotted.  Note that both axes in these plots
use a log scale.  In order to ensure a good sample of scalability we used 20
evenly spaced problem sizes $\left [ 5 .. 100 \right ]$ at increments of 5.
This information was then used to calculate each algorithms' order of complexity
in regards to problem size, with results given in Table~\ref{table-growth}.
A complete package of our results as well as the source code used in testing
is available from our website.\footnote{\url{https://name.of.site.removed.com/anonymous-CGP}}

@@

\section{Benchmarking}
\label{sec-benchmarking}
Understanding length biases and search limitations allows users to
solve new search problems easier.  In order to examine how our predicted search biases
interact with standard CGP landscapes, we tested all three algorithms on the 3-bit
multiplier problem.  In this problem there are six inputs and six outputs with
a function set of \{AND, OR, NAND, NOR\}.  Instead of scaling problem difficulty
as in Section~\ref{sec-tailored}, each algorithm was tested with scaling genome
sizes.  We used the range $\left [ 10 .. 8000 \right ]$, taking 21 points on an
exponential scale.  Other run parameters were kept the same.
The median number of evaluations to success is shown in
Figure~\ref{fig-multiply} on a log-log scale.  The results from the best genome
size are given in Table~\ref{table-multiply}.

As predicted, both the graphic and tabular results show that \emph{Reorder} and
\emph{DAG} can perform effective search with much smaller genome sizes than \emph{Normal}:
The best tested genome size for both novel
variants had lower median evaluations to success than \emph{Normal}'s best, which
also required a genome twenty times larger.  \emph{Reorder} was statistically
better than \emph{Normal} under the single tailed Mann-Whitney U test. While \emph{DAG} was not
statistically better than \emph{Normal}, the likely reason is \emph{Normal}'s exceptionally large median
absolute deviation.  We believe this deviation is caused by \emph{Normal}'s occasional need to
reinvent useful structures in different genome locations.  In the unlikely event that
active nodes are evolved with the right amount of inactive nodes between them, and in the
right order relative to each other, \emph{Normal} is likely going to receive far better
results.  In cases were these properties do not hold, \emph{Normal} will
have to recreate nodes in new locations, often with no fitness gradient, significantly
increasing run time.  As the number of nodes in the genome increases, the probability
that inactive nodes exist between two active nodes increases.  This may be part of
the reason \emph{Normal} requires such a large genome size to perform well.
Also, more inactive space in the genome may facilitate structure recreation in new locations.
As a final advantage to \emph{Reorder} and \emph{DAG}, we see that both evolved
solutions with half as many median active nodes as \emph{Normal}.  A potential
reason for this is \emph{Normal}'s need to leverage introns in order to perform
neutral search more than the other methods.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
	  \hline
& \textbf{\emph{Normal}} & \textbf{\emph{Reorder}} & \textbf{\emph{DAG}} \\ \hline
\textbf{Genome} & 4000   & 200     & 200 \\ \hline
  \textbf{MES} & 272,324 & 213,660 & 249,031 \\ \hline
  \textbf{MAD} & 125,997 & 79,335  & 86,551 \\ \hline
  \textbf{P value} & N/A & 0.030   & 0.117 \\ \hline
  \textbf{Active} & 255  & 118     & 120 \\ \hline

@@

\section{Conclusions and Future Work}
\label{sec-conclusion}
Existing literature suggests CGP performs best when maximum genome sizes
far exceed actual solution size~\cite{miller:2006:redundancy}, supported by
evidence that inactive genes improve neutral search~\cite{yu:2001:neutrality},
and that CGP doesn't suffer from bloat~\cite{miller:2001:bloat}.  We, however, find evidence
that the need for highly inactive genomes is likely due to limitations in CGP's
search mechanisms.  When no selective pressure on size exists, CGP highly favors
short individuals, with this search bias continuing even when the problem structure
calls for longer individuals.  Furthermore, when combined with CGP's use of node
positioning, problems requiring significant proportions of the genotype to be
active were found to be difficult or impossible for CGP to solve.  We showed theoretical
and experimental evidence that the blame for this problem lies with the way CGP's mutation
interacts with its representation, as existing mechanisms impose artificial search limitations
not inherent to the search problem.  \emph{Reorder}, which shuffled node positions
without changing node expression, and \emph{DAG}, which removed all concept of node
position, were both able to avoid these limitations and more fully utilize
genome space.  As these methods were able to achieve better results with less
inactive genes, we conclude that while inactive genes are useful, the number required
by \emph{Normal} CGP is a result of the inherent length bias and search limitations in its
representation and evolutionary mechanisms, and not because that many inactive
nodes are needed for neutral search.  Similarly, CGP's resistance to bloat is likely
no more than the work of a previously unexamined form of parsimony pressure, the strength
of which is determined by the maximum number of nodes.  The drawback of \emph{Reorder} and
\emph{DAG} is that in removing this parsimony pressure, they are more able to utilize
all genes in the genome, making them more sensitive to the genome size.  Yet this
may prove to be an advantage, as it may allow CGP to solve more complex problems without
explosively increasing the genome size.

Further analysis of \emph{Reorder} and \emph{DAG} should be performed in light
of the evidence found here.  Both appear to be valid ways of making CGP more flexible
and powerful, but wider testing needs to be performed to solidify that claim.  As
an important aspect of those tests, run time comparisons need to be performed, as
both increase the computational expense of CGP, with \emph{DAG} especially having
the potential for significant overhead.  It is also possible either could benefit
from further modification, such as using different initialization or method for mutation.

>>>>

\documentclass{sig-alternate}
%\linespread{2.5}
\usepackage{url}
\usepackage{verbatim}
\newcommand{\includegraphicsfit}[1]
{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{GECCO'13,} {July 6-10, 2013, Amsterdam, The Netherlands.}
\CopyrightYear{2013}
\crdata{TBA}
\clubpenalty=10000
\widowpenalty = 10000
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

\numberofauthors{2} %
%\begin{comment}
\author{
% 1st. author
\alignauthor
Brian W. Goldman\\
       \affaddr{BEACON Center for the Study of Evolution in Action}\\
       \affaddr{Michigan State University, U.S.A.}\\
       \email{brianwgoldman@acm.org}
% 2nd. author
\alignauthor
William F. Punch\\
       \affaddr{BEACON Center for the Study of Evolution in Action}\\
       \affaddr{Michigan State University, U.S.A.}\\
       \email{punch@msu.edu}
}
%\end{comment}
\begin{comment}
\author{
% 1st. author
\alignauthor
Anonymous\\
       \affaddr{Group}\\
       \affaddr{Organization}\\
       \email{email@site.com}
% 2nd. author
\alignauthor
Anonymous\\
       \affaddr{Group}\\
       \affaddr{Organization}\\
       \email{email@site.com}
}
%\end{comment}

@@

% A category with the (minimum) three required fields
%\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
\category{I.2.8}{Artificial Intelligence}{Problem Solving, Control Methods, and Search}
\category{I.2.2}{Artificial Intelligence}{Automatic Programming}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

@@

\keywords{Cartesian Genetic Programming, Neutrality, Bloat}

\section{Introduction}
In all black box search, search designers must make assumptions about the properties
of the problem in order to outperform random search.  When these assumptions
match the application, search time can be greatly reduced.  As such, understanding how design decisions
effect search effectiveness or even prevent search from finding solutions is very important.
%
In Cartesian Genetic Programming (CGP) the maximum genome size of an individual
is set at initialization.  While this
requires the maximum genome size to be larger than the problem's solution needs,
the question remains; how will the maximum genome size effect evolution's
speed at finding solutions.  As larger genomes inherently represent larger search
spaces, intuition suggests that using the lowest possible maximum size will have the most efficient search.
However, in CGP a mounting body of evidence suggests the optimal
maximum size is much larger than the actual solutions evolved~\cite{miller:2006:redundancy}.
We examine the cause of this discrepancy by applying CGP and a few
modified versions of CGP to problems which test specific edge cases in representation.
The result is a better understanding of the search limitations of CGP.

@@

For all problems except \emph{Neutral} (Section~\ref{sec-neutral}), we provide a plot of the median
evaluations until success using different problem sizes.  If the median run
was unsuccessful, no information is plotted.  Note that both axes in these plots
use a log scale.  In order to ensure a good sample of scalability we used 20
evenly spaced problem sizes $\left [ 5 .. 100 \right ]$ at increments of 5.
This information was then used to calculate each algorithms' order of complexity
in regards to problem size, with results given in Table~\ref{table-growth}.
A complete package of our results as well as the source code used in testing
is available from our website.\footnote{\url{https://github.com/brianwgoldman/LengthBiasCGP}}

@@

\section{Benchmarking}
\label{sec-benchmarking}
Understanding length biases and search limitations allows users to
solve new search problems easier.  In order to examine how our predicted search biases
interact with standard CGP landscapes, we tested all three algorithms on the 3-bit
multiplier problem.  In this problem there are six inputs and six outputs with
a function set of \{AND, OR, NAND, NOR\}.  Instead of scaling problem difficulty
as in Section~\ref{sec-tailored}, each algorithm was tested with scaling genome
sizes.  We used the range $\left [ 10 .. 8000 \right ]$, taking 21 points on an
exponential scale.  Other run parameters were kept the same.
The median number of evaluations to success is shown in
Figure~\ref{fig-multiply} on a log-log scale.  The runs using the best genome size for each 
algorithm were then repeated to allow for statistical comparison free from selection bias.
The results from the repeated runs are given in Table~\ref{table-multiply}.

As predicted, both the graphic and tabular results show that \emph{Reorder} and
\emph{DAG} can perform effective search with much smaller genome sizes than \emph{Normal}:
The best tested genome size for both novel
variants had lower median evaluations to success than \emph{Normal}'s best, which
also required a genome twenty times larger.  A Kruskal-Wallis Test was first used
to determine that at least one of the three algorithms had a different median than the others,
with a resulting p-value of 0.00006.
Both \emph{Reorder} and \emph{DAG} were statistically better than \emph{Normal} under
the single tailed Mann-Whitney~U test, even after accounting for multiple comparisons
using the Bonferroni correction.
Beyond the difference in median evaluations to success, \emph{Normal} also had
a much higher median absolute deviation.
We believe this deviation is caused by \emph{Normal}'s occasional need to
reinvent useful structures in different genome locations.  In the unlikely event that
active nodes are evolved with the right amount of inactive nodes between them, and in the
right order relative to each other, \emph{Normal} is likely going to receive far better
results.  In cases were these properties do not hold, \emph{Normal} will
have to recreate nodes in new locations, often with no fitness gradient, significantly
increasing run time.  As the number of nodes in the genome increases, the probability
that inactive nodes exist between two active nodes increases.  This may be part of
the reason \emph{Normal} requires such a large genome size to perform well.
Also, more inactive space in the genome may facilitate structure recreation in new locations.
The higher number of evaluations, higher deviation, and larger optimal genome size
also combined to make \emph{Normal} take almost three times more computational time
than either of the other algorithms to complete all runs.
As a final advantage to \emph{Reorder} and \emph{DAG}, we see that both evolved
solutions with half as many median active nodes as \emph{Normal}.  A potential
reason for this is \emph{Normal}'s need to leverage introns in order to perform
neutral search more than the other methods.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
	  \hline
& \textbf{\emph{Normal}} & \textbf{\emph{Reorder}} & \textbf{\emph{DAG}} \\ \hline
\textbf{Genome} & 4000   & 200     & 200 \\ \hline
  \textbf{MES} & 467,621 & 236,438 & 265,997 \\ \hline
  \textbf{MAD} & 215,993 & 82,966  & 66,930 \\ \hline
  \textbf{P-value} & N/A & 0.00012 & 0.00006 \\ \hline
  \textbf{Active} & 245  & 115     & 116 \\ \hline

@@

\section{Conclusions and Future Work}
\label{sec-conclusion}
Existing literature suggests CGP performs best when the maximum genome size
far exceeds the actual solution size~\cite{miller:2006:redundancy}, supported by
evidence that inactive genes improve neutral search~\cite{yu:2001:neutrality},
and that CGP doesn't suffer from bloat~\cite{miller:2001:bloat}.  We, however, find evidence
that the need for highly inactive genomes is likely due to limitations in CGP's
search mechanisms.  When no selective pressure on size exists, CGP highly favors
short individuals, with this search bias continuing even when the problem structure
calls for longer individuals.  Furthermore, when combined with CGP's use of node
positioning, problems requiring significant proportions of the genotype to be
active were found to be difficult or impossible for CGP to solve.  We showed theoretical
and experimental evidence that the blame for this problem lies with the way CGP's mutation
interacts with its representation, as existing mechanisms impose artificial search limitations
not inherent to the search problem.  \emph{Reorder}, which shuffled node positions
without changing node expression, and \emph{DAG}, which removed all concept of node
position, were both able to avoid these limitations and more fully utilize
genome space.  As these methods were able to achieve better results with less
inactive genes, we conclude that while inactive genes are useful, the number required
by \emph{Normal} CGP is a result of the inherent length bias and search limitations in its
representation and evolutionary mechanisms, and not because that many inactive
nodes are needed for neutral search.  Similarly, CGP's resistance to bloat is likely
no more than the work of a previously unexamined form of parsimony pressure, the strength
of which is determined by the maximum number of nodes.  The drawback of \emph{Reorder} and
\emph{DAG} is that in removing this parsimony pressure, they are more able to utilize
all genes in the genome, making them more sensitive to the genome size.  Yet this
may prove to be an advantage, as it may allow CGP to solve more complex problems without
explosively increasing the genome size.

Further analysis of \emph{Reorder} and \emph{DAG} should be performed in light
of the evidence found here.  Both appear to be valid ways of making CGP more flexible
and powerful, but wider testing needs to be performed to solidify that claim.  As
an important aspect of those tests, run time comparisons using equal genome sizes need to be performed, as
both increase the computational expense of CGP, with \emph{DAG} especially having
the potential for significant overhead.  It is also possible either could benefit
from further modification, such as using different initialization or method for mutation.
